working title Compatibility system and stygma size of pollen recipient as main predictors of heterospecific pollen effect

### 2018-12-08 Manuscript_draft/Manuscript_draft.Rmd
Some edits

Recent studies have advanced in the ecological understanding of heterospecific pollen effect @morales2008 @ashman2013 @arceo2016. A general overview of foreign pollen arrival is that it can play an important role on species fitness but seems to be context dependent and not always produce a decrease in fitness @morales2008. Part of this unpredictability is due to the enormous variability of foreing pollen transferred in nature, where levels between 0 and 75 percent are seen, but most commonly values ranges between 0 and 20 percent of the total pollen load, being the generalist species the ones that receive greater loads of heterospecific pollen @bartomeus2008 @montgomery2012 @ashman2013 @fang2013. Although heterospecific pollen quantity is fundamental to understand the outcome of the interaction so is the different traits of both pollen donor and recipient. @ashman2013 postulated the first predictive framework for traits of heterospecific pollen effect, where different traits such as compatibility system and pollen size among others seems to be crucial to understand foreing pollen effect. Moreover, in @tong2016 an assymetric effect was shown in a crossing experiment between 6 species of the genus *Pedicularis* where the pollen of long styled species was able to grow the full length of the style on short styled species but not viceversa. Despite these recent caveats, we still lack empirical evidence to affirm what are the main traits that drive heterospecific pollen effect for both pollen donor and recipient at seed production level. Interestingly, to comprehend how these traits interact is also crucial to look at the phylogenetic relatedness of the species. There is a considerable amount of literature of crosses between close related species @brown2001 @arceo2016can @tong2016 but few works focused on heterospecific pollen of far related species. Although the effect of close related species is predicted to be greater @ashman2013 the presence of pollen of non related species on multiple species @arceo2016 and the higher chances to coexist with a species that has less niche overlap (Ref) make foreign pollen from far related species also an important subject of study in order to understand the importance of heterospecific pollen in natural systems. Notwithstanding, the effect of heterospecific pollen of far and close related species at community level remains to be explored.

@@

Interestingly, incompatibility system seems to play an important role in foreign pollen effect where species that are self incompatible would have stronger barriers towards heterospecific pollen than self compatible species.  The type of incompatibility, sporophytic or gametophytic is related with the place of pollen recognition where the former take place at the sitgma level and the latter occurs within the style, this last late acting pollen recognition mechanism is associated with greater negative effect (REF). Remarkably, there is a great variability in mating systems across populations @whitehead2018 and therefore predict an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018). Moreover, other traits such as number of pollen grains per flower and number of ovules have been tradittionally associated with the type of incompatibility system where species with higher pollen ovule ratios are predicted to be xenogamous and species with low pollen ovule ratios autogamous (REFS).

>>>>

Recent studies have advanced in the ecological understanding of heterospecific pollen effect @morales2008 @ashman2013 @arceo2016. A general overview of foreign pollen arrival is that it can play an important role on species fitness but seems to be context dependent and not always produce a decrease in fitness @morales2008. Part of this unpredictability is due to the enormous variability of foreing pollen transferred in nature, where levels between 0 and 75 percent are seen, but most commonly values ranges between 0 and 20 percent of the total pollen load @bartomeus2008 @montgomery2012 @ashman2013 @fang2013, being the generalist species the ones that receive greater loads of heterospecific pollen @fang2013. Surprisingly, this low ranges of heterospecific pollen have been shown to decrease fitness greatly @thomson1982. Although heterospecific pollen quantity is fundamental to understand the outcome of the interaction so is the different traits of both pollen donor and recipient. @ashman2013 postulated the first predictive framework for traits of heterospecific pollen effect, where different traits such as compatibility system and pollen size among others seems to be crucial to understand foreing pollen effect. Moreover, in @tong2016 an assymetric effect was shown in a crossing experiment between 6 species of the genus *Pedicularis* where the pollen of long styled species was able to grow the full length of the style on short styled species but not viceversa. Despite these recent caveats, we still lack empirical evidence to affirm what are the main traits that drive heterospecific pollen effect for both pollen donor and recipient at seed production level. Interestingly, to comprehend how these traits interact is also crucial to look at the phylogenetic relatedness of the species. There is a considerable amount of literature of crosses between close related species @brown2001 @arceo2016can @tong2016 but few works focused on heterospecific pollen of far related species. Although the effect of close related species is predicted to be greater @ashman2013 the presence of pollen of non related species on multiple species @arceo2016 and the higher chances to coexist with a species that has less niche overlap (Ref) make foreign pollen from far related species also an important subject of study in order to understand the importance of heterospecific pollen in natural systems. Notwithstanding, the effect of heterospecific pollen of far and close related species at community level remains to be explored beyond single pairwise interactions.

@@

Interestingly, incompatibility system seems to play an important role in foreign pollen effect where species that are self incompatible would have stronger barriers towards heterospecific pollen than self compatible species @ashman2013. The type of incompatibility, sporophytic or gametophytic is related with the place of pollen recognition where the former take place at the sitgma level and the latter occurs within the style, this last late acting pollen recognition mechanism is associated with greater negative effect @barrett1999. Remarkably, there is a great variability in mating systems across populations @whitehead2018 and therefore predict an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018). Moreover, other traits such as number of pollen grains per flower and number of ovules have been tradittionally associated with the type of incompatibility system where species with higher pollen ovule ratios are predicted to be xenogamous and species with low pollen ovule ratios autogamous (REF). Selfer species would have a reduction of herkogamy (REF), less pollen production per ovule (REF) and therefore less likely to contribute in exporting pollen in the community. Other morphological traits, like stigma size can be determinant for the total pollen quatity that a stigma can receive and therefore related to do that pollen size would also play an important role. Example with pollen here.

### 2018-12-08 Manuscript_draft/Manuscript_draft.tex
Some edits

Recent studies have advanced in the ecological understanding of
heterospecific pollen effect Morales and Traveset (2008) Ashman and
Arceo-Gómez (2013) Arceo-Gómez and Ashman (2016). A general overview of
foreign pollen arrival is that it can play an important role on species
fitness but seems to be context dependent and not always produce a
decrease in fitness Morales and Traveset (2008). Part of this
unpredictability is due to the enormous variability of foreing pollen
transferred in nature, where levels between 0 and 75 percent are seen,
but most commonly values ranges between 0 and 20 percent of the total
pollen load, being the generalist species the ones that receive greater
loads of heterospecific pollen Bartomeus et al. (2008) Montgomery and
Rathcke (2012) Ashman and Arceo-Gómez (2013) Fang and Huang (2013).
Although heterospecific pollen quantity is fundamental to understand the
outcome of the interaction so is the different traits of both pollen
donor and recipient. Ashman and Arceo-Gómez (2013) postulated the first
predictive framework for traits of heterospecific pollen effect, where
different traits such as compatibility system and pollen size among
others seems to be crucial to understand foreing pollen effect.
Moreover, in Tong and Huang (2016) an assymetric effect was shown in a
crossing experiment between 6 species of the genus \emph{Pedicularis}
where the pollen of long styled species was able to grow the full length
of the style on short styled species but not viceversa. Despite these
recent caveats, we still lack empirical evidence to affirm what are the
main traits that drive heterospecific pollen effect for both pollen
donor and recipient at seed production level. Interestingly, to
comprehend how these traits interact is also crucial to look at the
phylogenetic relatedness of the species. There is a considerable amount
of literature of crosses between close related species Brown and
Mitchell (2001) Arceo-Gómez et al. (2016) Tong and Huang (2016) but few
works focused on heterospecific pollen of far related species. Although
the effect of close related species is predicted to be greater Ashman
and Arceo-Gómez (2013) the presence of pollen of non related species on
multiple species Arceo-Gómez and Ashman (2016) and the higher chances to
coexist with a species that has less niche overlap (Ref) make foreign
pollen from far related species also an important subject of study in
order to understand the importance of heterospecific pollen in natural
systems. Notwithstanding, the effect of heterospecific pollen of far and
close related species at community level remains to be explored.

@@

Interestingly, incompatibility system seems to play an important role in
foreign pollen effect where species that are self incompatible would
have stronger barriers towards heterospecific pollen than self
compatible species. The type of incompatibility, sporophytic or
gametophytic is related with the place of pollen recognition where the
former take place at the sitgma level and the latter occurs within the
style, this last late acting pollen recognition mechanism is associated
with greater negative effect (REF). Remarkably, there is a great
variability in mating systems across populations Whitehead et al. (2018)
and therefore predict an effect of foreign pollen is a bit obscured by
the variability within species, however species that are strong selfers
or strong outcrossers have less variablity in mating systems and
predictions of effect could be more realistic (see figure 1 from
Whitehead et al. (2018)).

@@

>>>>

Recent studies have advanced in the ecological understanding of
heterospecific pollen effect Morales and Traveset (2008) Ashman and
Arceo-Gómez (2013) Arceo-Gómez and Ashman (2016). A general overview of
foreign pollen arrival is that it can play an important role on species
fitness but seems to be context dependent and not always produce a
decrease in fitness Morales and Traveset (2008). Part of this
unpredictability is due to the enormous variability of foreing pollen
transferred in nature, where levels between 0 and 75 percent are seen,
but most commonly values ranges between 0 and 20 percent of the total
pollen load Bartomeus et al. (2008) Montgomery and Rathcke (2012) Ashman
and Arceo-Gómez (2013) Fang and Huang (2013), being the generalist
species the ones that receive greater loads of heterospecific pollen
Fang and Huang (2013). Surprisingly, this low ranges of heterospecific
pollen have been shown to decrease fitness greatly Thomson et al.
(1982). Although heterospecific pollen quantity is fundamental to
understand the outcome of the interaction so is the different traits of
both pollen donor and recipient. Ashman and Arceo-Gómez (2013)
postulated the first predictive framework for traits of heterospecific
pollen effect, where different traits such as compatibility system and
pollen size among others seems to be crucial to understand foreing
pollen effect. Moreover, in Tong and Huang (2016) an assymetric effect
was shown in a crossing experiment between 6 species of the genus
\emph{Pedicularis} where the pollen of long styled species was able to
grow the full length of the style on short styled species but not
viceversa. Despite these recent caveats, we still lack empirical
evidence to affirm what are the main traits that drive heterospecific
pollen effect for both pollen donor and recipient at seed production
level. Interestingly, to comprehend how these traits interact is also
crucial to look at the phylogenetic relatedness of the species. There is
a considerable amount of literature of crosses between close related
species Brown and Mitchell (2001) Arceo-Gómez et al. (2016) Tong and
Huang (2016) but few works focused on heterospecific pollen of far
related species. Although the effect of close related species is
predicted to be greater Ashman and Arceo-Gómez (2013) the presence of
pollen of non related species on multiple species Arceo-Gómez and Ashman
(2016) and the higher chances to coexist with a species that has less
niche overlap (Ref) make foreign pollen from far related species also an
important subject of study in order to understand the importance of
heterospecific pollen in natural systems. Notwithstanding, the effect of
heterospecific pollen of far and close related species at community
level remains to be explored beyond single pairwise interactions.

@@

Interestingly, incompatibility system seems to play an important role in
foreign pollen effect where species that are self incompatible would
have stronger barriers towards heterospecific pollen than self
compatible species Ashman and Arceo-Gómez (2013). The type of
incompatibility, sporophytic or gametophytic is related with the place
of pollen recognition where the former take place at the sitgma level
and the latter occurs within the style, this last late acting pollen
recognition mechanism is associated with greater negative effect
({\textbf{???}}). Remarkably, there is a great variability in mating
systems across populations Whitehead et al. (2018) and therefore predict
an effect of foreign pollen is a bit obscured by the variability within
species, however species that are strong selfers or strong outcrossers
have less variablity in mating systems and predictions of effect could
be more realistic (see figure 1 from Whitehead et al. (2018)). Moreover,
other traits such as number of pollen grains per flower and number of
ovules have been tradittionally associated with the type of
incompatibility system where species with higher pollen ovule ratios are
predicted to be xenogamous and species with low pollen ovule ratios
autogamous (REF). Selfer species would have a reduction of herkogamy
(REF), less pollen production per ovule (REF) and therefore less likely
to contribute in exporting pollen in the community. Other morphological
traits, like stigma size can be determinant for the total pollen quatity
that a stigma can receive and therefore related to do that pollen size
would also play an important role. Example with pollen here.

@@

\hypertarget{ref-thomson1982}{}
Thomson, J. D., B. J. Andrews, and R. Plowright. 1982. The effect of a
foreign pollen on ovule development in diervilla lonicera
(caprifoliaceae). New Phytologist 90:777--783.
